---
title: "Cell type prediction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cell type prediction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, warning=FALSE, message=FALSE}
library(cyCONDOR)
library(reticulate)
```

In this vignette we showcase how to use the `cyCONDOR` ecosystem to predict cell type and cell state without manual annotation of the dataset. This workflow is based on the `Astir` python package, if you use this workflow please consider citing the Astir manuscript [Geuenich et al. Cell Systems, 2021](https://www.sciencedirect.com/science/article/pii/S2405471221003355).

In `cyCONDOR` we use the `reticulate` package to run python code from R. If you use the `cyCONDOR` `Docker` image a conda environment is already configured to run `Astir`. If you have a local installation of `cyCONDOR` please visit `Astir` website to see a tutorial on how to get the tool installed in your system.

# Prepare your python environment

## Conda load
If you are using `cyCONDOR` `Docker` container you can visualize the list of available `conda` environment. 

```{r}
conda_list()
```
## Activate Conda

Now you simple need to activate `Astir` environment to be ready to run this workflow.

```{r}
use_condaenv(condaenv = "astir")
```

## Load example condor object

For this workflow we use an example dataset which was already analysed with `cyCONDOR`.

```{r}
condor <- readRDS("../.test_files/Astir/condor_example_astir.rds")
```

# Run Astir prediction

`Astir` allows to predict either cell type or cell state, for more details on the package see the official manuscript ([Geuenich et al. Cell Systems, 2021](https://www.sciencedirect.com/science/article/pii/S2405471221003355)).

For the prediction `Astir` needs a `manifest` file where the characteristinc of each cell type or cell state are specified.

This `manifest` file should be save as `.yml` file with this structure:

```{bash}
head ../.test_files/Astir/marker.yml -n 200
```

You can now run the two functions for the prediction of the cell type (`run_astir_celltype`) and cell state (`run_astir_cellstate`).

## Run `Astrir` to predict cell type

This function predict the cell type based on the marker selection specified in the manifest file. The output of this function is saved within the `condor` object under `condor$astir$Astir_cell_type_[data_slot]`. Additionally some QC data is saved in the `analysis_path` directory as `.csv`

```{r}
condor <- runAstir_celltype(fcd = condor, 
                            data_slot = "orig",
                            analysis_path = "../.test_files/Astir/",
                            manifest_name = "marker.yml", 
                            max_epochs = 1000, 
                            learning_rate = 0.002, 
                            initial_epochs = 3)
```

## Run `Astrir` to predict cell state

Similarly to the previous function the `runAstir_cellstates` function calculates a score for each cell state declared in the manifest file. The results are saved in the `condor` object in the slot `condor$astir$Astir_cell_state_[data_slot]`. Also in this case additional information are stored in the `analysis_path` directory in `.csv` format.

```{r}
condor <- runAstir_cellstates(fcd = condor, 
                              data_slot = "orig",
                              analysis_path = "../.test_files/Astir/",
                              manifest_name = "marker.yml", 
                              max_epochs = 1000, 
                              learning_rate = 0.002, 
                              initial_epochs = 3)
```

# Explore `Astrir` output

## Cell type predictiom

```{r}
head(read.csv("../.test_files/Astir/cell_types.csv"))
```

```{r}
head(read.csv("../.test_files/Astir/probabilities.csv"))
```

## Cell State

```{r}
head(read.csv("../.test_files/Astir/cell_states.csv"))
```

# Session Info
```{r}
info <- sessionInfo()

info
```
