---
title: "Clustering and cell annotation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clustering and cell annotation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, warning=FALSE, message=FALSE}
library(cyCONDOR)
```

In this vignette we showcase some `cyCONDOR` functions for clustering and cell annotation

# Load example cyCONDOR object
```{r}
condor <- readRDS("../.test_files/condor_example.rds")
```

The `cyCONDOR` ecosystem provides different clustering methods, Phenograph and FlowSOM. It also provide convenient functions to 

# Phenograph clustering
Phenograpg takes as input the `condor` object and the matrix to be used to the calculation of the clusters. We reccoment using the PCA coordinates as for non-linear dimensionality reduction. The `k` parameter define the number of nearest neighbors to be used for the nearest-neighbor graph. A seed can be set to ensure reproducibility of the clustering.

```{r}
condor <- runPhenograph(fcd = condor, 
                        input_type = "pca", 
                        data_slot = "orig", 
                        k = 60, 
                        seed = 91)
```

# FlowSOM clustering
FlowsOM provide a fast algorithm to cluster a high number of cells. The FlowSOM function takes the condor object as input and the matrix to be used to the calculation of the cluters. We reccoment using the PCA coordinates as for non-linear dimensionality reduction. The `num_cluster` parameter defines the final number of clusters to be generated. A seed can be set to ensure reproducibility of the clustering.
```{r}
condor <- runFlowSOM(fcd = condor, 
                     input_type = "pca", 
                     data_slot = "orig", 
                     num_clusters = 15, 
                     seed = 91, 
                     ret_model = TRUE)
```

# Metaclustering
Each cluster can be now labeled according to the specific cell type with the metaclustering function. This function takes as input the condor object and the clustering slot to be names. The name_out is the name of the new column containing the cell types. The metacluster parameter is a named vector where the content is the selected cell types and the names are the original cluster names.
```{r}
condor <- metaclustering(fcd = condor, 
                         clustering = "Phenograph_pca_orig_k_60", 
                         name_col = "Phenograph", 
                         name_out = "metaclusters", 
                         metaclusters = c("1" = "Classical Monocytes", 
                                          "2" = "B cells", 
                                          "3" = "CD8 CD45RA+ CD127-", 
                                          "4" = "CD8 CD45RA+ CD127-", 
                                          "5" = "NK bright", 
                                          "6" = "CD8 CD45RA- CD127+", 
                                          "7" = "CD4 CD45RA+ CD127+", 
                                          "8" = "CD8 CD45RA+ CD127+", 
                                          "9" = "CD4 CD127-", 
                                          "10" = "CD16- Monocytes", 
                                          "11" = "CD4 CD45RA- CD127+", 
                                          "12" = "Unconventional T cells", 
                                          "13" = "NK dim",
                                          "14" = "Classical Monocytes",
                                          "15" = "pDCs",
                                          "16" = "Classical Monocytes",
                                          "17" = "Classical Monocytes",
                                          "18" = "CD4 CD25+",
                                          "19" = "Unconventional T cells",
                                          "20" = "CD8 CD45RA- CD127+",
                                          "21" = "Basophils",
                                          "22" = "CD8 CD45RA- CD127+",
                                          "23" = "NK bright"))
```

# Session Info
```{r}
info <- sessionInfo()

info
```

