---
title: "condor_workflow - v0.1.3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{condor_workflow}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(condor)
```

# Loading the data

With this function you import the fcs files in R, the .fcs files should
be all stored in a single folder *FCSpath*, and the filename should
match one of the column names of the annotation file.

**Arguments:**

*`FCSpath`*= folder where the .fcs files are stored

*`ceil`*= Number of cells to use for each file (set to a high number if
you want to use all available events)

*`useCSV`* = flag if the input are .csv files and not .fcs
(experimental)

*`remove_param`* = Parameters to remove from the transformation,
"inTime" should be kept

*`anno_table`* = path to the annotation table file "filename_col" = Name
of the column containing the filename matching with the .fcs files

*`seed`*= seed to be used for the randomization of the events

```{r}
condor <- prep_fcd(FCSpath = "../.test_files/fcs/", 
                    ceil = 1000, 
                    useCSV = FALSE, 
                    transformation = "a", 
                    remove_param = c("FSC-H", "SSC-H", "FSC-W", "SSC-W", "Time", "InFile"), 
                    anno_table = "../.test_files/metadata.csv", 
                    filename_col = "filename",
                    seed = 91)

class(condor)
```

# Batch Correction

Batch correction can be performed with Harmony both at the level of the
transformed intensities and at level of PCA, we suggest to correct the
PCA but we give the possibility to correct also at the level of
intensities.

This function is adding to the flow cytometry dataframe the harmonized
intensities.

**Arguments:**

*`fcd`*= flow cytometry dataset

*`batch`*= vector of column names to use for correcting the data

*`seed`*= Seed used for the randomization steps

```{r}
condor <- harmonize_intensities(fcd = condor, 
                                batch = c("batch"), 
                                seed = 91)
```

# Dimensionality Reduction

## PCA

Can be calculated on the original data:

**Arguments:**

*`fcd`* = flow cytometry dataset

*`data_slot`* = name of the data slot to use to calculate the PCA,
original data (orig) or harmonized data (norm)

*`seed`* = Seed used for the randomization steps

```{r}
condor <- runPCA(fcd = condor, 
                 data_slot = "orig", 
                 seed = 91)
```

On this PCA you can run Harmony to remove batch effects, this is the
preferable method compared to correcting the expression table.

**Arguments:**

*`fcd`* = flow cytometry dataset

*`data_slot`* = name of the PCA data slot to use to harmonize. If no
prefix was added the, *orig*.

*`seed`* = Seed used for the randomization steps

```{r}
condor <- harmonize_PCA(fcd = condor, 
                        batch = c("batch"), 
                        data_slot = "orig", 
                        seed = 91)
```

## UMAP

On the original PCA

**Arguments:**

*`fcd`* = flow cytometry dataset

*`input type`* = data to use for the calculation of the UMAP, e.g.
"expr" or "pca"

*`data_slot`* = name of the PCA data slot to use to harmonize. If no
prefix was added the, *orig*.

*`seed`* = Seed used for the randomization steps

*`top_PCA`* = Select the number of PC to use.

```{r}
condor <- runUMAP(fcd = condor, 
                  input_type = "pca", 
                  data_slot = "orig", 
                  seed = 91)
```

On the harmonized PCA

```{r}
condor <- runUMAP(fcd = condor, 
                  input_type = "pca", 
                  data_slot = "norm", 
                  seed = 91, top_PCA = 3)
```

## tSNE

**Arguments:**

*`fcd`* = flow cytometry dataset

*`input type`* = data to use for the calculation of the UMAP, e.g.
"expr" or "pca"

*`data_slot`* = name of the PCA data slot to use to harmonize. If no
prefix was added the, *orig*.

*`seed`* = Seed used for the randomization steps

*`perplexity`* = Perplexity parameter for the calculation.

```{r}
condor <- runtSNE(fcd = condor, 
                  input_type = "pca", 
                  data_slot = "orig", 
                  seed = 91, 
                  perplexity = 30)
```


## Diffusion Map

**Arguments:**

*`fcd`* = flow cytometry dataset

*`input type`* = data to use for the calculation of the UMAP, e.g. "pca"
(suggested option)

*`data_slot`* = name of the PCA data slot to use to harmonize. If no
prefix was added the, *orig*.

*`seed`* = Seed used for the randomization steps

```{r}
condor <- runDM(fcd = condor, 
                input_type = "pca", 
                data_slot = "norm", 
                k = 10, 
                seed = 91)
```

# Clustering

## Phenograph

**Arguments:**

*`fcd`* = flow cytometry dataset.

*`input type`* = data to use for the calculation of the UMAP, e.g. "pca"
(suggested option).

*`data_slot`* = name of the PCA data slot to use to harmonize. If no
prefix was added the, *orig*.

*`k`* = K value used for clustering.

*`seed`* = Seed used for the randomization steps.

```{r}
condor <- runPhenograph(fcd = condor, 
                        input_type = "pca", 
                        data_slot = "orig", 
                        k = 60, 
                        seed = 91)
```

## FlowSOM

**Arguments:**

*`fcd`* = flow cytometry dataset.

*`input type`* = data to use for the calculation of the UMAP, e.g. "pca"
(suggested option).

*`data_slot`* = name of the PCA data slot to use to harmonize. If no
prefix was added the, *orig*.

*`num_clusters`* = Number of final clusters.

*`seed`* = Seed used for the randomization steps.

```{r}
condor <- runFlowSOM(fcd = condor, 
                     input_type = "expr", 
                     data_slot = "orig", 
                     num_clusters = 5, 
                     seed = 91, 
                     prefix = NULL, 
                     ret_model = TRUE)
```

### Specific visulaization for the FlowSOM model

```{r}
FlowSOM::PlotStars(condor$extras$FlowSOM_model)

FlowSOM::PlotMarker(condor$extras$FlowSOM_model, "CD14")

FlowSOM::PlotPies(condor$extras$FlowSOM_model, cellTypes = condor$clustering$Phenograph_pca_orig_k_60$Phenograph)
```



## Metaclustering

According to the cluster often you want to give a name or merge them
according to the marker expression. This is a function to help in this
task.

**Arguments:**

*`fcd`* = flow cytometry dataset.

*`clustering`* = Name of the clustering to match for the metaclustering.

*`name_col`* = Column containing the original cluster

*`name_out`* = Name of the output column

*`metaclusters`* = Vector of the new clusters names, this should be of
the same length of the levels of the original clustering.

As sanity check the function outputs a table with the matching between
the original clusters and the metaclusters to make it easy to check.

```{r}
condor <- metaclustering(fcd = condor, 
                         clustering = "Phenograph_pca_orig_k_60", 
                         name_col = "Phenograph", 
                         name_out = "metaclusters", 
                         metaclusters = c("1" = "CD4 T", 
                                          "2" = "B cells", 
                                          "3" = "CD8 T", 
                                          "4" = "Monocytes", 
                                          "5" = "CD8 T", 
                                          "6" = "CD8 T", 
                                          "7" = "Monocytes", 
                                          "8" = "Monocytes", 
                                          "9" = "Monocytes", 
                                          "10" = "CD4 T", 
                                          "11" = "CD4 T", 
                                          "12" = "NK cells", 
                                          "13" = "CD4 T"))
```

```{r}
condor <- metaclustering(fcd = condor, 
                         clustering = "FlowSOM_expr_orig_k_5", 
                         name_col = "FlowSOM", 
                         name_out = "metaclusters", 
                         metaclusters = c("1" = "CD4 T", 
                                          "2" = "NK cells", 
                                          "3" = "B cells", 
                                          "4" = "CD8 T", 
                                          "5" = "Monocytes"))
```

# Data visualization and statistics

## Dimensionality reduction visualization

Here a few examples, the function should be quite versatile.

**Arguments:**

*`data`* = This is the input dataframe for the visualization, this part
of the code can still be improved but for the moment you have to cbind
the dataframe with the information you want to plot.

*`param`* = Parameter to visualize in the plot, this can be either a
continuous variable or a categorical one, the function will react
differently according.

*`order`* = Logical if you want to order the dots in the plot, by
expression for example. This can help to find small populations of
positive cells.

*`title`* = Title of the plot.

*`dim_red`* = Dimensionality reduction to use for the visualization.

*`facet_by_variable`* = Logical if the plot should be split by the
categorical variable used.

```{r}
plot_marker(data = cbind(condor$umap$pca_orig, condor$clustering$Phenograph_pca_orig_k_60), 
            param = "Phenograph", 
            order = T, 
            title = "Phenograph Clustering on Normalized expression", 
            dim_red = "UMAP", 
            facet_by_variable = FALSE)

plot_marker(data = cbind(condor$umap$pca_orig, condor$clustering$Phenograph_pca_orig_k_60), 
            param = "Phenograph", 
            order = T, 
            title = "Phenograph Clustering on Normalized expression", 
            dim_red = "UMAP", 
            facet_by_variable = "metaclusters")

plot_marker(data = cbind(condor$umap$pca_orig, condor$clustering$Phenograph_pca_orig_k_60), 
            param = "Phenograph", 
            order = T, 
            title = "Phenograph Clustering on Normalized expression", 
            dim_red = "UMAP", 
            facet_by_variable = TRUE)
```


```{r}
plot_marker(data = cbind(condor$umap$pca_norm_top3, condor$clustering$Phenograph_pca_orig_k_60), 
            param = "Phenograph", 
            order = T, 
            title = "Phenograph Clustering on Normalized expression", 
            dim_red = "UMAP", 
            facet_by_variable = FALSE)

plot_marker(data = cbind(condor$tSNE$pca_orig, condor$clustering$Phenograph_pca_orig_k_60), 
            param = "Phenograph", 
            order = T, 
            title = "Phenograph Clustering on Normalized expression", 
            dim_red = "tSNE", 
            facet_by_variable = FALSE, dot_size = 2)
```

```{r}
plot_marker(data = cbind(condor$umap$pca_orig, condor$clustering$Phenograph_pca_orig_k_60), 
            param = "Phenograph", 
            order = T, 
            title = "Phenograph Clustering on Normalized expression", 
            dim_red = "UMAP", 
            facet_by_variable = FALSE)
```

```{r}
plot_marker(data = cbind(condor$umap$pca_orig, condor$clustering$FlowSOM_expr_orig_k_5), 
            param = "FlowSOM", 
            order = T, 
            title = "Phenograph Clustering on Normalized expression", 
            dim_red = "UMAP", 
            facet_by_variable = FALSE)
```

```{r}
plot_marker(data = cbind(condor$umap$pca_orig, condor$clustering$Phenograph_pca_orig_k_60), 
            param = "metaclusters", 
            order = T, 
            title = "Phenograph Clustering on Normalized expression", 
            dim_red = "UMAP", facet_by_variable = FALSE)

plot_marker(data = cbind(condor$umap$pca_orig, condor$clustering$Phenograph_pca_orig_k_60, condor$anno$cell_anno),
            param = "metaclusters",
            order = T,
            title = "Phenograph Clustering on Normalized expression",
            dim_red = "UMAP", facet_by_variable = "group")

plot_marker(data = cbind(condor$umap$pca_orig, condor$clustering$Phenograph_pca_orig_k_60, condor$anno$cell_anno), 
            param = "metaclusters", 
            order = T, 
            title = "Phenograph Clustering on Normalized expression", 
            dim_red = "UMAP", facet_by_variable = FALSE)

plot_marker(data = cbind(condor$umap$pca_orig, condor$clustering$Phenograph_pca_orig_k_60, condor$anno$cell_anno), 
            param = "Phenograph", 
            order = T, 
            title = "Phenograph Clustering on Normalized expression", 
            dim_red = "UMAP", facet_by_variable = FALSE, label_clusters = TRUE)
```


```{r, fig.width=15}
plot_density(data = cbind(condor$umap$pca_orig, condor$clustering$Phenograph_pca_orig_k_60, condor$anno$cell_anno), 
             param = "group", 
             title = "test", 
             dim_red = "UMAP", 
             dot_size = 0.2, 
             alpha = 0.2)
```

```{r}
plot_marker(data = cbind(condor$umap$pca_orig, condor$clustering$FlowSOM_expr_orig_k_5), 
            param = "metaclusters", 
            order = T, 
            title = "Phenograph Clustering on Normalized expression", 
            dim_red = "UMAP", facet_by_variable = FALSE, label_clusters = TRUE)
```

```{r}
for (marker in colnames(condor$expr$orig)[3:29]) {
  
  plot(plot_marker(data = cbind(condor$umap$pca_orig, condor$expr$orig), 
                   param = marker, 
                   order = T, 
                   title = "Phenograph Clustering on Original expression", 
                   dim_red = "UMAP", facet_by_variable = FALSE))
  
}
```

## Confusion Matrix

**Arguments:**

*`variable`* = Variable to by used to calculate the confusion.

*`group`* = Grouping to calculate the relative contribution to the
variable.

*`title`* = Title for the plot.

```{r}
confusion_HM(variables = condor$clustering$Phenograph_pca_orig_k_60$Phenograph, 
             group = condor$anno$cell_anno$group, size = 30, 
             title = "Confusion Matrix - Phenograph K60")
```

```{r}
confusion_HM(variables = condor$clustering$Phenograph_pca_orig_k_60$metaclusters, 
             group = condor$anno$cell_anno$group, size = 30, 
             title = "Confusion Matrix - Metaclusters")
```

## Barplot Frequencies

**Arguments:**

*`x_axes`* = Grouping of the x axes.

*`color`* = Stratification to use on the stacked barplot.

*`title`* = Title for the plot.

*`legend_title`* = Title for the legend.

```{r}
barplot_frequency(x_axes = condor$anno$cell_anno$group, 
                  colour = condor$clustering$Phenograph_pca_orig_k_60$Phenograph, 
                  title = "Stacked barplot of cluster frequencies", 
                  legend_title = "Phenograph Cluster - k60")
```

```{r}
barplot_frequency(x_axes = condor$anno$cell_anno$group, 
                  colour = condor$clustering$Phenograph_pca_orig_k_60$metaclusters, 
                  title = "Stacked barplot of metaclusters frequencies", 
                  legend_title = "Phenograph Cluster - Metaclusters")
```

## Expression HM

**Arguments:**

*`input`* = cbind of the expression table to be used the the grouping
(eg. clustering).

*`group`* = column name to be used for the grouping (eg. "Phenograph" or
"group")

*`maxvalue`* = Max scaled expression to be used for the color coding.

*`title`* = Title for the plot.

```{r, fig.height=7}
HM_markers(input = cbind(condor$expr$orig, condor$clustering$Phenograph_pca_orig_k_60), 
           group = "Phenograph",
           maxvalue = 2, 
           title = "Marker expression in the clusters", 
           cluster_rows = TRUE)
```

```{r, fig.height=7}
HM_markers(input = cbind(condor$expr$orig, condor$clustering$Phenograph_pca_orig_k_60), 
           group = "metaclusters",
           maxvalue = NULL, 
           title = "Marker expression in metaclusters")
```

## Boxplot and statistics

**Arguments:**

*`annotation`* = Sample annotation to be used for the plot.

*`sample_var`* = Column name containing the sample IDs.

*`group_var`* = Column name defining the grouping for plotting.

*`variable`* = Variable used to stratify the plotting.

*`numeric`* = Logical if the grouping is numeric.

*`test.type`* = Test to be performed. (see need some development here)

```{r}
box_stat <- boxplot_and_stats(annotation = condor$anno$cell_anno, 
                              sample_var = "sample_ID", 
                              group_var = "group", 
                              variable = condor$clustering$Phenograph_pca_orig_k_60$metaclusters, 
                              numeric = FALSE, 
                              test.type = "t.test")

box_stat
```

## Extra functions

### Check integrity of the flow cytometry dataframe

**Arguments:**

*`fcd`* = Flow cytometry dataset to be checked

```{r}
# This function checks the integrity of the  
check_IDs(condor)
```

### Subset the flow cytometry dataset

**Arguments:**

*`fcdataset`* = Flow cytometry dataset to be filtered.

*`cell_ids`* = Row names of the cells to be filtered, should be provided
as vector.


* This need to be fixed*
```{r}
selections <- rownames(condor$clustering$Phenograph_pca_orig_k_60[condor$clustering$Phenograph_pca_orig_k_60$metaclusters == "CD4 T", ])

condor_filter <- filter_fcd(fcdataset = condor,
                            cell_ids = selections)
```

```{r}
plot_marker(data = cbind(condor_filter$umap$pca_norm, condor_filter$clustering$Phenograph_pca_orig_k_60),
            param = "metaclusters",
            order = T,
            title = "Phenograph Clustering on Normalized expression",
            dim_red = "UMAP", facet_by_variable = FALSE)
```

# Session Info

```{r}
info <- sessionInfo()

info
```
