---
title: "cyCONDOR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cyCONDOR}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

# Getting Started

This is a basic `cyCONDOR` workflow for data analysis, more detailed
description of all `cyCONDOR` functionalists and visualization tools can
be found in the *Articles* section. In this section we will describe how
to load data from a folder of fcs files, how to perform dimensionality
reduction and clustering together with some basic visualization of the
results.

We start by loading the cyCONDOR package

```{r setup, warning=FALSE, message=FALSE}
library(cyCONDOR)
```

# Loading the data

With this function you import the `.fcs` files in R, the `.fcs` files should
be all stored in a single folder *FCSpath*, and the file names should
match one column of the annotation file.

**Arguments:**

*`FCSpath`*= folder where the `.fcs` files are stored

*`ceil`*= Number of cells to use for each file (set to a high number if
you want to use all available events)

*`useCSV`* = flag if the input are `.csv` files and not `.fcs`
(experimental)

*`remove_param`* = Parameters to remove from the transformation,
"inTime" should be kept

*`anno_table`* = path to the annotation table file "filename_col" = Name
of the column containing the filename matching with the .fcs files

*`seed`*= seed to be used for the randomization of the events

```{r}
condor <- prep_fcd(FCSpath = "../.test_files/fcs/", 
                    ceil = 1000, 
                    useCSV = FALSE, 
                    transformation = "a", 
                    remove_param = c("FSC-H", "SSC-H", "FSC-W", "SSC-W", "Time", "InFile"), 
                    anno_table = "../.test_files/metadata.csv", 
                    filename_col = "filename",
                    seed = 91)

class(condor)
```

# Dimensionality Reduction

To reduce the complexity of the dataset we perform first dimensionality reduction, we first perform a principal component analysis (PCA) and use those coordinates for non-linear dimensionality reduction with the UMAP or tSNE algorithms. With this approach we can visualize the complexity of the dataset in a two dimensional space.

## PCA

**Arguments:**

*`fcd`* = cyCONDOR dataset

*`data_slot`* = name of the data slot to use to calculate the PCA,
original data (orig) or harmonized data (norm)

*`seed`* = Seed used for the randomization steps

```{r}
condor <- runPCA(fcd = condor, 
                 data_slot = "orig", 
                 seed = 91)
```

## UMAP

**Arguments:**

*`fcd`* = cyCONDOR dataset

*`input type`* = data to use for the calculation of the UMAP, e.g.
"expr" or "pca"

*`data_slot`* = name of the PCA data slot to use to harmonize. If no
prefix was added the, *orig*.

*`seed`* = Seed used for the randomization steps

*`top_PCA`* = Select the number of PC to use.

```{r}
condor <- runUMAP(fcd = condor, 
                  input_type = "pca", 
                  data_slot = "orig", 
                  seed = 91)
```

### Visualize the UMAP 

```{r}
plot_marker(data = cbind(condor$umap$pca_orig, condor$anno$cell_anno), 
            param = "group", 
            order = T, 
            title = "UMAP colored by group", 
            dim_red = "UMAP", 
            facet_by_variable = FALSE)
```

## tSNE

**Arguments:**

*`fcd`* = cyCONDOR dataset

*`input type`* = data to use for the calculation of the UMAP, e.g.
"expr" or "pca"

*`data_slot`* = name of the PCA data slot to use to harmonize. If no
prefix was added the, *orig*.

*`seed`* = Seed used for the randomization steps

*`perplexity`* = Perplexity parameter for the calculation.

```{r}
condor <- runtSNE(fcd = condor, 
                  input_type = "pca", 
                  data_slot = "orig", 
                  seed = 91, 
                  perplexity = 30)
```

### Visualize the tSNE

```{r}
plot_marker(data = cbind(condor$tSNE$pca_orig, condor$anno$cell_anno), 
            param = "group", 
            order = T, 
            title = "tSNE colored by group", 
            dim_red = "tSNE", 
            facet_by_variable = FALSE)
```

# Clustering

We now group cells with similar marker expression with the Phenograph or FlowSOM clustering algorithms. 

## Phenograph

**Arguments:**

*`fcd`* = cyCONDOR dataset.

*`input type`* = data to use for the calculation of the UMAP, e.g. "pca"
(suggested option).

*`data_slot`* = name of the PCA data slot to use to harmonize. If no
prefix was added the, *orig*.

*`k`* = K value used for clustering.

*`seed`* = Seed used for the randomization steps.

```{r}
condor <- runPhenograph(fcd = condor, 
                        input_type = "pca", 
                        data_slot = "orig", 
                        k = 60, 
                        seed = 91)
```

### Visualize Phenograph results

```{r}
plot_marker(data = cbind(condor$umap$pca_orig, condor$clustering$Phenograph_pca_orig_k_60), 
            param = "Phenograph", 
            order = T, 
            title = "UMAP colored by Phenograph clustering", 
            dim_red = "UMAP", 
            facet_by_variable = FALSE)
```

## FlowSOM

**Arguments:**

*`fcd`* = cyCONDOR dataset.

*`input type`* = data to use for the calculation of the UMAP, e.g. "pca"
(suggested option).

*`data_slot`* = name of the PCA data slot to use to harmonize. If no
prefix was added the, *orig*.

*`num_clusters`* = Number of final clusters.

*`seed`* = Seed used for the randomization steps.

```{r}
condor <- runFlowSOM(fcd = condor, 
                     input_type = "expr", 
                     data_slot = "orig", 
                     num_clusters = 5, 
                     seed = 91, 
                     prefix = NULL, 
                     ret_model = TRUE)
```

### Visualize FlowSOM results

```{r}
plot_marker(data = cbind(condor$umap$pca_orig, condor$clustering$FlowSOM_expr_orig_k_5), 
            param = "FlowSOM", 
            order = T, 
            title = "UMAP colored by Flowsom clustering", 
            dim_red = "UMAP", 
            facet_by_variable = FALSE)
```

# Data visualization and statistics

We can now visualize the results or the analysi and compare different experimental groups.

## Confusion Matrix

**Arguments:**

*`variable`* = Variable to by used to calculate the confusion.

*`group`* = Grouping to calculate the relative contribution to the
variable.

*`title`* = Title for the plot.

```{r}
confusion_HM(variables = condor$clustering$Phenograph_pca_orig_k_60$Phenograph, 
             group = condor$anno$cell_anno$group, size = 30, 
             title = "Confusion Matrix - Metaclusters")
```

## Barplot Frequencies

**Arguments:**

*`x_axes`* = Grouping of the x axes.

*`color`* = Stratification to use on the stacked barplot.

*`title`* = Title for the plot.

*`legend_title`* = Title for the legend.

```{r}
barplot_frequency(x_axes = condor$anno$cell_anno$group, 
                  colour = condor$clustering$Phenograph_pca_orig_k_60$Phenograph, 
                  title = "Stacked barplot of metaclusters frequencies", 
                  legend_title = "Phenograph Cluster - Metaclusters")
```

## Expression HM

**Arguments:**

*`input`* = cbind of the expression table to be used the the grouping
(eg. clustering).

*`group`* = column name to be used for the grouping (eg. "Phenograph" or
"group")

*`maxvalue`* = Max scaled expression to be used for the color coding.

*`title`* = Title for the plot.

```{r, fig.height=5}
HM_markers(input = cbind(condor$expr$orig, condor$clustering$Phenograph_pca_orig_k_60), 
           group = "Phenograph",
           maxvalue = NULL, 
           title = "Marker expression in metaclusters")
```

## Boxplot and statistics

**Arguments:**

*`annotation`* = Sample annotation to be used for the plot.

*`sample_var`* = Column name containing the sample IDs.

*`group_var`* = Column name defining the grouping for plotting.

*`variable`* = Variable used to stratify the plotting.

*`numeric`* = Logical if the grouping is numeric.

*`test.type`* = Test to be performed. (see need some development here)

```{r}
box_stat <- boxplot_and_stats(annotation = condor$anno$cell_anno, 
                              sample_var = "sample_ID", 
                              group_var = "group", 
                              variable = condor$clustering$Phenograph_pca_orig_k_60$Phenograph, 
                              numeric = FALSE, 
                              test.type = "t.test")

box_stat$plot$Cluster_7
```

# Session Info

```{r}
info <- sessionInfo()

info
```
