---
title: "Data Projection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Projection}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, warning=FALSE, message=FALSE}
library(cyCONDOR)
```

Considering the high nunber of samples that can be generated with modern HDC instruments and the number of cells acquired we developed a data projection workflow in `cyCONDOR`. With this approach an initial model it trained including a UMAP dimensionality reduction and a cell classifier of the cell labels. We can then project any new sample into the pre-trained model, this operation is much faster and allows to analyse millions of cell in few minutes. 

# Loading the data for training

We start by loading the data for the training.

```{r}
condor_train <- prep_fcd(FCSpath = "../.test_files/fcs/", 
                        ceil = 5000, 
                        useCSV = FALSE, 
                        transformation = "a", 
                        remove_param = c("FSC-H", "SSC-H", "FSC-W", "SSC-W", "Time", "InFile", "live_dead"), 
                        anno_table = "../../../Figure 6 - Data Projection/data_and_envs/metadata_train.csv", 
                        filename_col = "filename",
                        seed = 91)
```
```{r}
condor_train$anno$cell_anno$group <- "train"
```

# Loading the data for projection

We also load the data to later project.

```{r}
condor_test <- prep_fcd(FCSpath = "../.test_files/fcs/", 
                        ceil = 5000, 
                        useCSV = FALSE, 
                        transformation = "a", 
                        remove_param = c("FSC-H", "SSC-H", "FSC-W", "SSC-W", "Time", "InFile", "live_dead"), 
                        anno_table = "../../../Figure 6 - Data Projection/data_and_envs/metadata_test.csv", 
                        filename_col = "filename",
                        seed = 91)
```

```{r}
condor_test$anno$cell_anno$group <- "test"
```

# UMAP Projection

We start now by running a UMAP, in this case we set the `ret_model` variable to `TRUE` to keep the UMAP model in the `condor`  object.

## Run UMAP keeping the model
```{r}
condor_train <- runUMAP(fcd = condor_train, 
                        input_type = "expr", 
                        data_slot = "orig", 
                        seed = 91, 
                        n_threads = 4, 
                        ret_model = TRUE)
```

## Add data to the embedding

We can now predict the UMAP coordinates of the projected data.

```{r}
condor_test <- learnUMAP(fcd = condor_test, 
                         input_type = "expr", 
                         data_slot = "orig", 
                         model = condor_train$extras$umap_model, 
                         n_epochs = 100, 
                         seed = 91, 
                         n_threads = 4)
```

# Train a classifier for the label transfer

To transfer also the labels from the reference to the projected data we need to train a cell classifier. We start by clustering the training data.

```{r}
condor_train <- runPhenograph(fcd = condor_train, 
                              input_type = "expr", 
                              data_slot = "orig", 
                              k = 300, 
                              seed = 91)
```

## Visualization
```{r}
plot_marker(data = cbind(condor_train$umap$expr_orig, condor_train$clustering$Phenograph_expr_orig_k_300), 
            param = "Phenograph", 
            order = T, 
            title = "Figure S9b -UMAP on the training dataset, Phenograph clusters", 
            dim_red = "UMAP", 
            facet_by_variable = FALSE, 
            label_clusters = FALSE)
```

# Label transfer

And we now train the classifier on the clustering labels. If you assigned a metacluster label, this can also be used to train the classifier.

## Train label transfer kNN classifier
```{r}
condor_train_fine <- train_transfer_model(fcd = condor_train, 
                                          data_slot = "orig", 
                                          input_type = "expr", 
                                          label = condor_train$clustering$Phenograph_expr_orig_k_300$Phenograph, 
                                          method = "knn", 
                                          tuneLength = 5, 
                                          trControl = caret::trainControl(method = "cv"), 
                                          seed = 91)
```

```{r}
condor_train_fine$extras$lt_model$performace_plot
```

```{r, fig.height=10, fig.width=10}
# Figure S10b - kNN importance

condor_train_fine$extras$lt_model$features_plot
```

## Predict the labels

```{r}
condor_test <- predict_labels(fcd = condor_test, 
                              data_slot = "orig", 
                              input_type = "expr", 
                              model_object = condor_train_fine, 
                              label = "label_pred_fine", 
                              seed = 91)
```

# Visualize the results

## Prepare the dataframe
```{r}
train <- as.data.frame(cbind(condor_train$umap$expr_orig, Phenograph = condor_train$clustering$Phenograph_expr_orig_k_300[, 1]))

train$type <- "train"

test <- cbind(condor_test$umap$expr_orig, 
              condor_test$clustering$label_pred_fine)

test$Description <- NULL
test$Description <- NULL

colnames(test) <- c("UMAP1", "UMAP2", "Phenograph")

test$type <- "test"

vis_data <- rbind(train, test)
```

# Overlap UMAP
```{r}
vis_data$type <- factor(vis_data$type, levels = c("train", "test"))
```

```{r}
ggplot(data = vis_data, aes(x = UMAP1, y = UMAP2, color = type, alpha = type, size = type)) +
  geom_point() +
  scale_color_manual(values = c("gray", "#92278F")) +
  scale_alpha_manual(values = c(0.5, 1)) +
  scale_size_manual(values = c(0.1, 0.5)) +
  theme_bw() +
  theme(aspect.ratio = 1, panel.grid = element_blank()) + 
  ggtitle("Figure 6c - UMAP projected")
```

```{r}
ggplot(data = vis_data, aes(x = UMAP1, y = UMAP2, color = Phenograph, alpha = type, size = type)) +
  geom_point() +
  scale_alpha_manual(values = c(0.1, 1)) +
  scale_size_manual(values = c(0.1, 0.5)) +
  theme_bw() +
  theme(aspect.ratio = 1, panel.grid = element_blank()) + 
  ggtitle("Figure S9c - Predicted cluster")
```

# Session Info
```{r}
info <- sessionInfo()

info
```

