---
title: "Data Visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Visualization}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, warning=FALSE, message=FALSE}
library(cyCONDOR)
```

cyCONDOR provides several function to visualize the result of the analysis, following some examples on how to use them.

# Load an example dataset
```{r}
condor <- readRDS("../.test_files/condor_example.rds")
```

# Plot the results of dimensionality reduction and clustering

## PCA
```{r}
plot_marker(data = cbind(condor$pca$orig, condor$anno$cell_anno), 
            param = "group", 
            order = T, 
            title = "UMAP colored by group", 
            dim_red = "PCA", 
            facet_by_variable = FALSE)
```

## tSNE
```{r}
plot_marker(data = cbind(condor$tSNE$pca_orig, condor$anno$cell_anno), 
            param = "group", 
            order = T, 
            title = "tSNE colored by group", 
            dim_red = "tSNE", 
            facet_by_variable = FALSE)
```

## UMAP
```{r}
plot_marker(data = cbind(condor$umap$pca_orig, condor$anno$cell_anno), 
            param = "group", 
            order = T, 
            title = "UMAP colored by group", 
            dim_red = "UMAP", 
            facet_by_variable = FALSE)
```

## Colored by Phenograph, FlowSOM clustering or metaclusters
```{r}
plot_marker(data = cbind(condor$umap$pca_orig, condor$anno$cell_anno, condor$clustering$Phenograph_pca_orig_k_60), 
            param = "Phenograph", 
            order = T, 
            title = "UMAP colored by Phenograph clusters", 
            dim_red = "UMAP", 
            facet_by_variable = FALSE)
```

```{r}
plot_marker(data = cbind(condor$umap$pca_orig, condor$anno$cell_anno, condor$clustering$FlowSOM_pca_orig_k_15), 
            param = "FlowSOM", 
            order = T, 
            title = "UMAP colored by FlowSOM clusters", 
            dim_red = "UMAP", 
            facet_by_variable = FALSE)
```

```{r}
plot_marker(data = cbind(condor$umap$pca_orig, condor$anno$cell_anno, condor$clustering$Phenograph_pca_orig_k_60), 
            param = "metaclusters", 
            order = T, 
            title = "UMAP colored by metaclusters", 
            dim_red = "UMAP", 
            facet_by_variable = FALSE)
```

## Export the plot as raster image
```{r}
plot_marker(data = cbind(condor$umap$pca_orig, condor$anno$cell_anno, condor$clustering$Phenograph_pca_orig_k_60), 
            param = "metaclusters", 
            order = T, 
            title = "UMAP colored by metaclusters", 
            dim_red = "UMAP", 
            facet_by_variable = FALSE, 
            raster = TRUE)
```

## Split plot by a variable
```{r}
plot_marker(data = cbind(condor$umap$pca_orig, condor$anno$cell_anno), 
            param = "group", 
            order = T, 
            title = "UMAP colored by group", 
            dim_red = "UMAP", 
            facet_by_variable = TRUE)
```

```{r}
plot_marker(data = cbind(condor$umap$pca_orig, condor$anno$cell_anno, condor$clustering$Phenograph_pca_orig_k_60), 
            param = "metaclusters", 
            order = T, 
            title = "UMAP colored by metaclusters", 
            dim_red = "UMAP", 
            facet_by_variable = "group")
```

# Density plot
```{r, fig.width=15}
plot_density(data = cbind(condor$umap$pca_orig, condor$anno$cell_anno), 
             param = "group", 
             title = "Figure S2d - Density Map", 
             dim_red = "UMAP", 
             dot_size = 0.2, 
             alpha = 0.2, color_density = c("Greys", "Reds"))
```


# Heatmap of marker expression
```{r, fig.width=7, fig.height=7}
HM_markers(input = cbind(condor$expr$orig, condor$clustering$Phenograph_pca_orig_k_60), 
           group = "metaclusters",
           maxvalue = NULL, 
           title = "Marker expression in metaclusters")
```

# Visualize the loadings of the PCs
```{r}
PC_loadings(fcd = condor, data_slot = "orig", number = 1)
```

# Violinplot of marker expression
```{r}
violinplot_marker(fcd = condor, 
                  data_slot = "orig", 
                  cluster_method = "Phenograph_pca_orig_k_60", 
                  cluster_type = "metaclusters", marker = "CD56")
```

# Classical cytometry dotplot
```{r}
dotplot_cyto(data = cbind(condor$expr$orig, condor$clustering$Phenograph_pca_orig_k_60), 
             subset = TRUE,
             annotation = condor$clustering$Phenograph_pca_orig_k_60$metaclusters, 
             subset_char = c("NK bright", "NK dim"), 
             color_by = condor$clustering$Phenograph_pca_orig_k_60$metacluster, 
             x = "CD16", 
             y = "CD56", 
             title = "NK cells Dotplot")
```

# Densityplot marker
```{r}
densityplot_marker(data = cbind(condor$expr$orig, condor$clustering$Phenograph_pca_orig_k_60), 
                   marker = "CD3", 
                   split_by = NULL, 
                   color_by = "metaclusters", 
                   title = "CD3 expression")
```

# Session Info
```{r}
info <- sessionInfo()

info
```

