---
title: "Differential Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differential Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, warning=FALSE, message=FALSE}
library(cyCONDOR)
library(ggpubr)
```

# Load an example dataset
```{r}
condor <- readRDS("../.test_files/condor_diff.rds")
```

`cyCONDOR` provides several ways to investigate the difference in cellular frequencies or marker expression.

# Cell Frqeuency

## Export Perfecntage table
If you prefer to generate your own plots, `cyCONDOR` can easily export a percentage table for downstream analysis.
```{r}
df_frequency(classification = condor$clustering$Phenograph_pca_orig_k_60$metaclusters, 
             condition = condor$anno$cell_anno$group)
```

## Confusion Matrix
```{r, fig.width=15, fig.height=5}
confusion_HM(variables = condor$clustering$Phenograph_pca_orig_k_60$metaclusters, 
             group = condor$anno$cell_anno$group, size = 40, 
             title = "Confusion Matrix")
```

# Stacked barplot
```{r}
barplot_frequency(x_axes = condor$anno$cell_anno$group, 
                  colour = condor$clustering$Phenograph_pca_orig_k_60$metaclusters, 
                  title = "Stacked barplot of cluster frequencies", 
                  legend_title = "Metaclusters") + theme(aspect.ratio = 2)
```

## Percentages Boxplot 
```{r, fig.width=20, fig.height=20}
box_stat <- boxplot_and_stats(annotation = condor$anno$cell_anno, 
                              sample_var = "sample_ID", 
                              group_var = "group", 
                              variable = condor$clustering$Phenograph_pca_orig_k_60$metaclusters, 
                              numeric = FALSE, 
                              test.type = "t.test")

ggarrange(plotlist = box_stat$plot, legend = "none")
```

# Differential expression

## Splitted heatmap by group
```{r, fig.width=20, fig.height=20}
HM_differential_marker(fcd = condor, 
                       data_slot = "orig", 
                       cluster_method = "Phenograph_pca_orig_k_60", 
                       cluster_type = "metaclusters", 
                       group_by = "group",
                       title = "Figure 5f - Marker expression per cluster and group", size = 40)
```

# Session Info
```{r}
info <- sessionInfo()

info
```

