---
title: "Dimensionality Reduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dimensionality Reduction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, warning=FALSE, message=FALSE}
library(cyCONDOR)
```

In this vignette we showcase some `cyCONDOR` functions for dimensionality reduction. Here we exemplify the how to run PCA, UMAP. tSNE and DM. All functions take as input the `fcd` and the data slot to be used for the calculation. The `runPCA` function always use the `expr` slot for the calculation, for non linear dimensionality reduction (UMAP, tSNE, DM) the used can decide to use the `expr` data or the `pca` result as input.

# Load example cyCONDOR object
```{r}
condor <- readRDS("../.test_files/condor_example.rds")
```

# Principal Component Analysis
```{r}
condor <- runPCA(fcd = condor, 
                 data_slot = "orig", 
                 seed = 91)
```

# tSNE
```{r}
condor <- runtSNE(fcd = condor, 
                  input_type = "pca", 
                  data_slot = "orig", 
                  seed = 91, 
                  perplexity = 30)
```

# UMAP
```{r}
condor <- runUMAP(fcd = condor, 
                  input_type = "pca", 
                  data_slot = "orig", 
                  seed = 91)
```

# Diffusion Map
```{r}
condor <- runDM(fcd = subset_fcd(condor, 5000), 
                input_type = "pca", 
                data_slot = "orig", 
                k = 10, 
                seed = 91)
```

# Session Info
```{r}
info <- sessionInfo()

info
```

