---
title: "Machine learning classifier"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Machine learning classifier}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, warning=FALSE, message=FALSE}
library(cyCONDOR)
library(ggplot2)
library(CytoDx)
```

With `cyCONDOR` we developed a set of functions allowing the user to easily train a classifier for the sample labels. For example we can train the model on a dataset of control and leukemia sample. The trained model can then be used to predict the label of and external sample. This workflow is based on the `CytoDX` package

# Loading the data - Training Set

We start by importing the training dataset:

```{r}
condor <- prep_fcd(FCSpath = "../../../Figure 7 - Clinical Classifier/data_and_envs/CytoDX/", 
                    ceil = 10000000, 
                    useCSV = FALSE, 
                    transformation = "a", 
                    remove_param = c("FSC-A","FSC-W","FSC-H","Time", "InFile"), 
                    anno_table = "../../../Figure 7 - Clinical Classifier/data_and_envs/CytoDX/fcs_info_train.csv", 
                    filename_col = "fcsName",
                    seed = 91)
```

## Build the classifier model

We now train the classifier on the sample label, this step does not require any other pre-analysis on the dataset, nevertheless, if you are not familiar with the data you are using from training we recommend an exploratory data analysis firts.

```{r}
# Re order variables

condor$anno$cell_anno$Label <- factor(condor$anno$cell_anno$Label, levels = c("normal", "aml"), labels = c("1_normal", "2_aml"))
```

```{r}
condor <- train_classifier_model(fcd = condor, 
                                 input_type = "expr", 
                                 data_slot = "orig", 
                                 sample_names = "expfcs_filename", 
                                 classification_variable = condor$anno$cell_anno$Label, 
                                 family = "binomial", 
                                 type1 = "response", 
                                 parallelCore = 1, 
                                 reg = FALSE, 
                                 seed = 91)
```

### Explore the model

We can now explore the results of the model at sample level

```{r}
head(condor$extras$classifier_model$train.Data.sample)
```

And at cellular level

```{r}
head(condor$extras$classifier_model$train.Data.cell)
```


# Load the data - test Set

To now validate the performance of the trained model we will test it on a test dataset with no overlap with the training data.

```{r}
condor_test <- prep_fcd(FCSpath = "../../../Figure 7 - Clinical Classifier/data_and_envs/CytoDX/", 
                    ceil = 10000000, 
                    useCSV = FALSE, 
                    transformation = "a", 
                    remove_param = c("FSC-A","FSC-W","FSC-H","Time", "InFile"), 
                    anno_table = "../../../Figure 7 - Clinical Classifier/data_and_envs/CytoDX/fcs_info_test.csv", 
                    filename_col = "fcsName",
                    seed = 91)
```

## Predict classification

We can now predict the label using the trained model

```{r}
# Re order variables

condor_test$anno$cell_anno$Label <- factor(condor_test$anno$cell_anno$Label, 
                                           levels = c("normal", "aml"), labels = c("1_normal", "2_aml"))
```

```{r}
condor_test <- predict_classifier(fcd = condor_test, 
                                  input_type = "expr", 
                                  data_slot = "orig", 
                                  sample_names = "expfcs_filename", 
                                  model_object = condor$extras$classifier_model, 
                                  seed = 91)
```

### Explore Prediction

Also here we can explore the results at sample level

```{r}
head(condor_test$extras$classifier_prediction$xNew.Pred.sample)
```

And at cellular level

```{r}
head(condor_test$extras$classifier_prediction$xNew.Pred.cell)
```

# Visualize the results

## Training Data
```{r}
anno <- read.csv("../../../Figure 7 - Clinical Classifier/data_and_envs/CytoDX/fcs_info_train.csv")

tmp <- merge(x = condor$extras$classifier_model$train.Data.cell, y = anno, by.x = "sample", by.y = "fcsName")

ggplot(tmp, aes(x = sample, y = y.Pred.s0, color = Label)) +
  geom_jitter() + 
  geom_violin() +
  scale_color_manual(values = c("#92278F", "#F15A29")) +
  theme_bw() + 
  theme(aspect.ratio = 1) + 
  ggtitle("Figure S12b - Training prediction cell")
```

```{r}
tmp <- merge(x = condor$extras$classifier_model$train.Data.sample, y = anno, by.x = "sample", by.y = "fcsName")

ggplot(tmp, aes(x = sample, y = y.Pred.s0, color = Label)) +
  geom_point(size = 4) +
  scale_color_manual(values = c("#92278F", "#F15A29")) +
  theme_bw() + 
  theme(aspect.ratio = 2) + 
  ggtitle("Figure S12c - Training prediction sample")
```

## Test Data
```{r}
anno <- read.csv("../../../Figure 7 - Clinical Classifier/data_and_envs/CytoDX/fcs_info_test.csv")

tmp <- merge(x = condor_test$extras$classifier_prediction$xNew.Pred.cell, y = anno, by.x = "sample", by.y = "fcsName")

tmp$sample <- factor(tmp$sample, levels = c("sample1.fcs", "sample2.fcs", "sample3.fcs",  "sample4.fcs",  "sample5.fcs",  "sample6.fcs",  "sample7.fcs",  "sample8.fcs",  "sample9.fcs", "sample10.fcs"))

ggplot(tmp, aes(x = sample, y = y.Pred.s0, color = Label)) +
  geom_jitter() +
  geom_violin() +
  scale_color_manual(values = c("#92278F", "#F15A29")) +
  theme_bw() + 
  theme(aspect.ratio = 1) + 
  ggtitle("Figure 7c - Test prediction cell")
```

```{r}
tmp <- merge(x = condor_test$extras$classifier_prediction$xNew.Pred.sample, y = anno, by.x = "sample", by.y = "fcsName")

tmp$sample <- factor(tmp$sample, levels = c("sample1.fcs", "sample2.fcs", "sample3.fcs",  "sample4.fcs",  "sample5.fcs",  "sample6.fcs",  "sample7.fcs",  "sample8.fcs",  "sample9.fcs", "sample10.fcs"))

ggplot(tmp, aes(x = sample, y = y.Pred.s0, color = Label)) +
  geom_point(size = 4) +
  scale_color_manual(values = c("#92278F", "#F15A29")) +
  theme_bw() + 
  theme(aspect.ratio = 2) + 
  ggtitle("Figure 7d - Test prediction sample")
```

# Visualization of the decision three
```{r}
treeGate(P = condor$extras$classifier_model$train.Data.cell$y.Pred.s0,
         x= condor$expr$orig)
```

# Session Info
```{r}
info <- sessionInfo()

info
```
