---
title: "Other utilities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Other utilities}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, warning=FALSE, message=FALSE}
library(cyCONDOR)
```

In this vignette we showcase some `cyCONDOR` utilities

# Load example cyCONDOR object
```{r}
condor <- readRDS("../.test_files/condor_example.rds")
```

# Check the integrity of the condor object
This function can be useful to make sure the condor objet has the right structure for all downstream analysis
```{r}
check_IDs(condor)
```

# Change parameter names
This function can be used to change the name or the parameters in the condor object. In this example we chane the name from PD-1 to PD1
```{r}
old_names <- colnames(condor$expr$orig)

new_names <- old_names

new_names[22] <- "PD1"

condor_changed <- change_param_name(condor, old_names, new_names)
```

# Merge two condor object
With this function you can merge two `condor` objects, the two objects need to have the same parameters. This function will merge only expression table and annotation as all the downstream analysis will need to be repeated.
```{r}
condor_merged <- merge_condor(condor, condor)
```

# Subset a condor object
With this function you can subset a `condor` object to a specific number of cells.
```{r}
condor_subset <- subset_fcd(condor, 5000)
```

# Filter a condor object to create a specific subset
This function can be useful to created a specific subset of a `condor` object
```{r}
condor_filter <- filter_fcd(condor, 
                            cell_ids = rownames(condor$expr$orig)[condor$clustering$Phenograph_pca_orig_k_60$metaclusters == "Classical Monocytes"])
```

# Session Info
```{r}
info <- sessionInfo()

info
```
